---
title: "Beta Binomial approach for SNPs"
output: html_notebook
---

## Load libraries and clear out memory

```{r}
library(phyclust)
library(rstan)
library(tidyverse)
library(ggplot2)

rm(list = ls())

options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```

The stationary distribution of allele frequencies within a population given effective size $N_e$, mutation rate $\mu$ from $A_1$ to $A_2$, and mutation rate $\nu$ from $A_2$ to $A_1$ is

$$
\begin{align*}
P(p) &= \frac{1}{B(4N_e\nu - 1, 4N_e\mu - 1)}p^{4N_e\nu - 1}(1 - p)^{4N_e\mu - 1} \\
     &= \frac{1}{B(\phi - 1, \theta - 1)}p^{\phi - 1}(1 - p)^{\theta - 1}
\end{align*}
$$
where $\phi = \theta\frac{\nu}{\mu}$.

Consider a sample of SNPs from $L$ loci with $n_l$ the number of "1" alleles at locus $l$ in a sample of $N_l$ alleles.

$$
P(n_l | \phi, \theta) = \mbox{BetaBinomial}(n_l | N_l, \phi, \theta)
$$

With $L$ loci we have $L+1$ parameters to estimate, even if we assume $N_e$ is constant across loci. That means the likelihood is non-identifiable, and any estimate using this approach will be very sensitive to the choice of priors. If, however, we assume $\nu$ and $\mu$ are constant across loci, then we have only two parameters to estimate. Let's see how well that works, if at all.

```{r}
construct_n <- function(dna_seqs, n_sample) {
  n <- numeric(n_sample)
  if (length(seg.sites(dna_seqs)) > 0 ) {
    n_seqs <- nrow(dna_seqs)
    dna_seqs_chr <- character(n_seqs)
    for (i in 1:n_seqs) {
      dna_seqs_chr[i] <- paste(dna_seqs[i, ], collapse = "")
    }
    counts <- table(as.factor(dna_seqs_chr))
    for (count in 1:n_sample) {
      index <- counts[count]
      n[index] <- n[index] + 1
    }
  } else {
    n[n_sample] <- 1
  }
  return(which(n == 1)[1])
}

## Run the simulation for a finite length sequence
##
## Note: ts_tv = 0.5 gives equal rates since there are 4 types of transversions and only 
## 2 types of transitions
##
simulate_finite <- function(n_sample, seq_len, n_loci, theta,  
                            shape = 1, rate = 1, n_rep = 1, ts_tv = 0.5, equal_pars = FALSE) 
{
  phi_B <- numeric(n_rep)
  theta_B <- numeric(n_rep)
  for (rep in 1:n_rep) {
    if (n_rep > 1) {
      cat(".")
      if (rep %% 5 == 0) {
        count <- sprintf("%4d", rep)
        cat(count)
      }
      if (rep %% 50 == 0) {
        cat("\n")
      }
    }
    ## simulate the data
    ##
    n <- numeric(n_loci)
    for (l in 1:n_loci) {
      seg_sites <- 0
      ## if conditional == TRUE, keep constructing samples until we have a SNP
      ## Note: as written there could be more than two alleles at the SNP
      ##
      while ((seg_sites != 1)) {
        ms_tree <- phyclust::ms(nsam = n_sample, nreps = 1, opts = paste("-T"))
        ## N.B.: seqs_chr comes back in PHYLIP format as a vector of strings (including the taxon)
        ## HKY with -t0.5 and without -f (frequencies) makes the model Jukes-Cantor
        ##
        options <- paste("-mHKY -l", seq_len, " -d", theta, " -t", ts_tv, " -q", sep = "")
        seqs_chr <- seqgen(opts = options, newick.tree = ms_tree[3])
        ## We write the result to disk
        ##
        write(seqs_chr, "seqs.phy")
        ## read it back using read.dna (from ape) and delete it
        ##
        tmp <- read.dna("seqs.phy")
        file.remove("seqs.phy")
        ## check the number of segregating sites
        ##
        seg_sites <- length(seg.sites(tmp))
      }
      n[l] <- construct_n(tmp, n_sample)
    }
    ## Bayesian estimate: single theta shared across all loci
    ##
    stan_data <- list(N = n_sample,
                      L = n_loci,
                      n = n,
                      equal_pars = (equal_pars == TRUE)) 
    model <- "beta-binomial-SNP.stan"
    fit <- stan(file = model,
                data = stan_data,
                refresh = (n_rep == 1))
    fit_df <- as.data.frame(fit)
    phi_B[rep] <- mean(fit_df$phi, na.rm = TRUE)
    theta_B[rep] <- mean(fit_df$theta, na.rm = TRUE)
  }
  ret_val <- tibble(phi_B = phi_B,
                    theta_B = theta_B)
  return(ret_val)
}
```

```{r}
results <- simulate_finite(n_sample = 25, seq_len = 250, n_loci = 100, theta = 0.0001, n_rep = 10,
                           shape = 1, rate = 100, equal_pars = FALSE)

summary(results)

for_plot <- results %>%
  pivot_longer(everything())
p <- ggplot(for_plot, aes(x = value, color = name, fill = name)) +
  geom_density(alpha = 0.2) +
  geom_vline(xintercept = 0.0001, color = "salmon", linetype = "dashed") +
  theme_bw() +
  theme(axis.text.y = element_blank())
p
```


```{r}
results <- simulate_finite(n_sample = 25, seq_len = 250, n_loci = 100, theta = 0.0001, n_rep = 10,
                           shape = 1, rate = 100, equal_pars = TRUE)

summary(results)

for_plot <- results %>%
  pivot_longer(everything())
p <- ggplot(for_plot, aes(x = value, color = name, fill = name)) +
  geom_density(alpha = 0.2) +
  geom_vline(xintercept = 0.0001, color = "salmon", linetype = "dashed") +
  theme_bw() +
  theme(axis.text.y = element_blank())
p
```

