---
title: "Multilocus estimates of pi (infinite sites)"
output: html_notebook
---

# Overview

We'll simulate $L$ independently inherited SNPs, using an infinite sites model and examining only those samples that have only one segregating site. We'll estimate $\theta$ Ewen's sampling formula, $\theta_B$, where we assume the same $\theta_B$ across all loci. There's no point in trying $\theta_k$, since it will always give the same answer for a particular sample size.

## Load libraries and clear out memory

```{r}
library(phyclust)
library(rstan)
library(tidyverse)
library(ggplot2)

rm(list = ls())

options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```

## Utility functions

```{r}
## construct matrix of allele counts, loci in rows
##
construct_a <- function(ms_tree, n_sample, seg_sites) {
  a <- numeric(n_sample)
  if (seg_sites > 0) {
    counts <- table(ms_tree[5:(n_sample+4)])
    for (count in 1:length(counts)) {
      index <- counts[count]
      a[index] <- a[index] + 1
    }
  } else {
    a[n_sample] <- 1
  }
  return(a)
}

## Run the simulation for a finite length sequence
##
simulate_infinite <- function(n_sample, n_loci, theta, 
                              n_rep = 1, shape = 1, rate = 1) 
{
  theta_B <- numeric(n_rep)
  for (rep in 1:n_rep) {
    if (n_rep > 1) {
      cat(".")
      if (rep %% 5 == 0) {
        count <- sprintf("%4d", rep)
        cat(count)
      }
      if (rep %% 50 == 0) {
        cat("\n")
      }
    }
    ## simulate the data
    ##
    a <- matrix(nrow = n_loci, ncol = n_sample)
    for (l in 1:n_loci) {
      seg_sites <- -1
      while (seg_sites != 1) {
        ms_tree <- phyclust::ms(nsam = n_sample, nreps = 1, opts = paste("-t", theta))
        seg_sites <- as.numeric(gsub(".* ([0-9]+)", "\\1", ms_tree[3]))
      }
      ## construct a matrix for Bayesian estimate
      ##
      a[l, ] <- construct_a(ms_tree, n_sample, seg_sites)
    }
    ## Bayesian estimate: single theta shared across all loci
    ##
    stan_data <- list(N = n_sample,
                      L = n_loci,
                      a = a,
                      shape = shape,
                      rate = rate) 
    fit <- stan(file = "ewens-multilocus.stan",
                data = stan_data,
                refresh = 0)
    fit_df <- as.data.frame(fit)
    theta_B[rep] <- mean(fit_df$theta, na.rm = TRUE)
    #cat(theta_k[rep], " ", theta_B[rep], "\n")
  }
  ret_val <- tibble(theta_B = theta_B,
                    eps_theta_B = theta - theta_B)
  return(ret_val)
}
```

## Run a simulation

```{r}
results <- simulate_infinite(n_sample = 25,
                             n_loci = 100,
                             theta = 0.1,
                             n_rep = 10)

cat("theta_B: ", mean(results$theta_B), "(", sqrt(mean(results$eps_theta_B^2)), ")", "\n", sep = "")

for_plot <- pivot_longer(results, cols = everything(), names_to = "Model") %>%
  filter(Model == "theta_k" | Model == "theta_B")
p <- ggplot(for_plot, aes(x = value, fill = Model, color = Model)) +
  geom_density(alpha = 0.2) +
  geom_vline(xintercept = 0.1, color = "salmon", linetype = "dashed") +
  theme_bw()
p
```

Using SNPs with this approach is clearly a bad idea, unless maybe you simulate short sequences and place mutations with seqgen.