---
title: "Using Ewens sampling distribution to estimate &theta;"
output: html_notebook
---

## Load libraries and clear out memory

```{r}
library(phyclust)
library(rstan)
library(tidyverse)
library(ggplot2)

rm(list = ls())

options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```

## Define utility functions

```{r}
## calculate a harmonic sum: \sum_i 1/i
##
harm_sum <- function(n) {
  k <- 0
  for (i in 1:(n-1)) {
    k <- k + 1/i
  }
  return(k)
}

## Watterson's estimate of theta
##
est_theta <- function(seg_sites, n_sample) {
  return(seg_sites/harm_sum(n_sample))
}

## get mean and 95% credible intervals from Stan fit
##
get_fit_stats <- function(fit) {
  fit_dat <- as.data.frame(fit)
  mu <- mean(fit_dat$theta, na.rm = TRUE)
  lo <- quantile(fit_dat$theta, 0.025)
  hi <- quantile(fit_dat$theta, 0.975)
  return(list(mu = mu, lo = lo, hi = hi))
}

## Run the simulation
##
simulate <- function(n_sample, shape, rate, theta, n_rep = 1) {
  theta_watterson <- numeric(n_rep)
  theta_bayes <- tibble(rep = 0, theta = 0, lp__ = 0)[0, ]
  for (rep in 1:n_rep) {
    if (n_rep > 1) {
      cat(".")
      if (rep %% 5 == 0) {
        count <- sprintf("%4d", rep)
        cat(count)
      }
      if (rep %% 50 == 0) {
        cat("\n")
      }
    }
    ## simulate the data
    ##
    ms_tree <- phyclust::ms(nsam = n_sample, nreps = 1, opts = paste("-t", theta))
    ## estimate theta using Watterson approach
    ##
    seg_sites <- as.numeric(gsub(".* ([0-9]+)", "\\1", ms_tree[3]))
    theta_watterson[rep] <- est_theta(seg_sites, n_sample)
    ## estimate theta using Bayesian approach
    ##
    a <- numeric(n_sample)
    if (seg_sites > 0) {
      counts <- table(ms_tree[5:(n_sample+4)])
      for (count in 1:length(counts)) {
        index <- counts[count]
        a[index] <- a[index] + 1
      }
    } else {
      a[n_sample] <- 1
    }
    stan_data <- list(N = length(a),
                      a = a,
                      shape = shape,
                      rate = rate)
    fit <- stan(file = "ewens.stan",
                data = stan_data,
                refresh = 0)
    theta_bayes <- add_row(theta_bayes, cbind(rep = rep, as.data.frame(fit)))
  }
  return(list(theta_watterson = theta_watterson, theta_bayes = theta_bayes))
}
```

## Explore prior choices

```{r}
## prior on theta
##   mean = shape/rate
##   variance = shape/(rate^2)
##
p <- ggplot() + 
  xlim(0, 5) + 
  geom_function(fun = dgamma, args = list(shape = 1, rate = 1), color = "salmon") +
  geom_function(fun = dgamma, args = list(shape = 0.5, rate = 0.5), color = "royalblue") +
  geom_function(fun = dgamma, args = list(shape = 2, rate = 2), color = "orange") +
  geom_function(fun = dgamma, args = list(shape = 4, rate = 4), color = "blueviolet") +
  geom_text(aes(x = 4, y = 1.4), label = "shape = 1, rate = 1", color = "salmon") +
  geom_text(aes(x = 4, y = 1.5), label = "shape = 0.5, rate = 0.5", color = "royalblue") +
  geom_text(aes(x = 4, y = 1.3), label = "shape = 2, rate = 2", color = "orange") +
  geom_text(aes(x = 4, y = 1.2), label = "shape = 4, rate = 4", color = "blueviolet") +
  xlab(bquote(theta)) +
  ylab("") +
  theme_bw()
p

parameters <- matrix(c(1, 1,
                       0.5, 0.5,
                       2, 2,
                       4, 4), byrow = TRUE, ncol = 2)
```

## Compare prior and posterior for $\theta = 0.1$, `n_sample = 25`

```{r}
theta <- numeric(0)
shape <- numeric(0)
rate <- numeric(0)
parameter_set <- character(0)
for (i in 1:nrow(parameters)) {
  result <- simulate(n_sample = 25, shape = parameters[i, 1], rate = parameters[i, 2], 
                     theta = 0.1, n_rep = 1)
  result_dat <- result$theta_bayes
  n_in_sample <- nrow(result_dat)
  theta <- c(theta, result_dat$theta)
  shape <- c(shape, rep(parameters[i, 1], n_in_sample))
  rate <- c(rate, rep(parameters[i, 1], n_in_sample))
  tmp <- sprintf("shape = %3.1f, rate = %3.1f", parameters[i, 1], parameters[i, 2])
  parameter_set <- c(parameter_set, rep(tmp, n_in_sample))
}
fit_dat <- tibble(theta = theta,
                  shape = shape,
                  rate = rate,
                  parameter_set = parameter_set)

theta <- numeric(0)
density <- numeric(0)
parameter_set <- character(0)
x <- seq(0, 2, 0.01)
for (i in 1:nrow(parameters)) {
  y <- dgamma(x, shape = parameters[i, 1], rate = parameters[i, 2])
  theta <- c(theta, x)
  density <- c(density, y)
  tmp <- sprintf("shape = %3.1f, rate = %3.1f", parameters[i, 1], parameters[i, 2])
  parameter_set <- c(parameter_set, rep(tmp, length(x)))
}
dens_dat <- tibble(theta = theta,
                   density = density,
                   parameter_set = parameter_set)

## Plot posterior vs. prior
##
title <- paste("Sample size: 25, Prior density shown as dashed red line. ",
               bquote(theta), 
               "= 0.1.")
p <- ggplot(fit_dat, aes(x = theta)) +
  geom_density() +
  geom_line(data = dens_dat, aes(x = theta, y = density), color = "salmon", linetype = "dashed") +
  xlab(bquote(theta)) +
  ylab("") +
  facet_wrap(~ parameter_set) +
  ggtitle(title) +
  theme_bw()
p
```

## Compare prior and posterior for $\theta = 0.1$, `n_sample = 250`

```{r}
theta <- numeric(0)
shape <- numeric(0)
rate <- numeric(0)
parameter_set <- character(0)
for (i in 1:nrow(parameters)) {
  result <- simulate(n_sample = 250, shape = parameters[i, 1], rate = parameters[i, 2], 
                     theta = 0.1, n_rep = 1)
  result_dat <- result$theta_bayes
  n_in_sample <- nrow(result_dat)
  theta <- c(theta, result_dat$theta)
  shape <- c(shape, rep(parameters[i, 1], n_in_sample))
  rate <- c(rate, rep(parameters[i, 1], n_in_sample))
  tmp <- sprintf("shape = %3.1f, rate = %3.1f", parameters[i, 1], parameters[i, 2])
  parameter_set <- c(parameter_set, rep(tmp, n_in_sample))
}
fit_dat <- tibble(theta = theta,
                  shape = shape,
                  rate = rate,
                  parameter_set = parameter_set)

theta <- numeric(0)
density <- numeric(0)
parameter_set <- character(0)
x <- seq(0, 2, 0.01)
for (i in 1:nrow(parameters)) {
  y <- dgamma(x, shape = parameters[i, 1], rate = parameters[i, 2])
  theta <- c(theta, x)
  density <- c(density, y)
  tmp <- sprintf("shape = %3.1f, rate = %3.1f", parameters[i, 1], parameters[i, 2])
  parameter_set <- c(parameter_set, rep(tmp, length(x)))
}
dens_dat <- tibble(theta = theta,
                   density = density,
                   parameter_set = parameter_set)

## Plot posterior vs. prior
##
title <- paste("Sample size: 250, Prior density shown as dashed red line. ",
               bquote(theta), 
               "= 0.1.")
p <- ggplot(fit_dat, aes(x = theta)) +
  geom_density() +
  geom_line(data = dens_dat, aes(x = theta, y = density), color = "salmon", linetype = "dashed") +
  xlab(bquote(theta)) +
  ylab("") +
  facet_wrap(~ parameter_set) +
  ggtitle(title) +
  theme_bw()
p
```

## Set simulation parameters

$\Gamma(1, 1)$ seems like the best choice of prior (based on this very cursory study). The posterior is more clearly different from the prior in larger (250) samples. The weight of the density at 0 is also much less than with $\Gamma(0.5, 0.5)$. Let's see how the Watterson and Bayesian estimates compare using samples of 250 alleles across 100 replicates with $\theta = 0.1$ using a $\Gamma(1,1)$ prior on $\theta$.

```{r}
n_repetitions <- 100
theta <- 0.1

result <- simulate(n_sample = 250, shape = 1, rate = 1, theta = theta, n_rep = n_repetitions)
theta_watterson <- result$theta_watterson
theta_bayes <- numeric(0)
theta_lo <- numeric(0)
theta_hi <- numeric(0)
for (i in 1:n_repetitions) {
  rep_results <- subset(result$theta_bayes, rep == i)
  tmp <- get_fit_stats(rep_results)
  theta_bayes <- c(theta_bayes, tmp$mu)
  theta_lo <- c(theta_lo, tmp$lo)
  theta_hi <- c(theta_hi, tmp$hi)
}
dat <- tibble(theta_watterson = theta_watterson,
              theta_bayes = theta_bayes,
              theta_lo = theta_lo,
              theta_hi = theta_hi)
dat_sum <- dat %>%
  select(theta_watterson, theta_bayes) %>%
  summarize(theta_w = mean(theta_watterson),
            theta_w_sd = sd(theta_watterson),
            theta_b = mean(theta_bayes),
            theta_b_sd = sd(theta_bayes))
cover <- 0
for (i in 1:nrow(dat)) {
  if ((dat$theta_lo[i] < theta) && (dat$theta_hi[i] > theta)) {
    cover <- cover + 1
  }
}
theta_w_bias <- mean(dat$theta_watterson - theta)
theta_w_rmse <- sqrt(sum(dat$theta_watterson - theta)^2)
theta_b_bias <- mean(dat$theta_bayes - theta)
theta_b_rmse <- sqrt(sum(dat$theta_bayes - theta)^2)
dat_sum$theta_w_bias <- theta_w_bias
dat_sum$theta_w_rmse <- theta_w_rmse
dat_sum$theta_b_bias <- theta_b_bias
dat_sum$theta_b_rmse <- theta_b_rmse
dat_sum$theta_b_cover <- cover/n_repetitions
dat_sum

theta_df <- tibble(theta_watterson = theta, theta_bayes = theta)
p <- ggplot(dat, aes(x = theta_watterson, y = theta_bayes)) +
  geom_point() +
  geom_point(data = theta_df, size = 4, color = "royalblue") +
  geom_abline(slope = 1, intercept = 0, color = "salmon", linetype = "dashed") +
  xlim(0, NA) +
  ylim(0, NA) +
  ggtitle("Watterson vs. Bayesian estimates") +
  theme_bw()
p

dat <- dat %>% 
  select(theta_watterson, theta_bayes) %>% 
  pivot_longer(cols = everything(), names_to = "Model", values_to = "theta") %>%
  mutate(Model = substring(Model, 7))
p <-ggplot(dat, aes(x = theta)) +
  geom_histogram(binwidth = 0.05) +
  geom_vline(xintercept = theta, color = "salmon", linetype = "dashed") +
  facet_wrap(~ Model) +
  theme_bw()
p
```

